---
title: "Bar Graph Demo"
author: "Jens von Bergmann"
date: '2017-08-17'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bar Graph Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

We demonstrate how to import data from the [CensusMapper](https://CensusMapper.ca) API. The code for this notbook can be found on [GitHub](https://github.com/mountainMath/cancensus).

This example demonstrates how to dynamically discover regions and vectors and produce an overview graph based on the data by using the `list_census_regions`, `list_census_vectors` and `search_census_vectors` methods.

```{r API Key, echo=TRUE, warning=FALSE}
library(cancensus)
# set your API key
# options(cancensus.api_key='your_api_key')

# use 2016 data
dataset <- 'CA16'
```

### Dynamically Select Regions

We select the top 6 CMAs by population.
```{r, echo=TRUE, message=FALSE, warning=FALSE}
# grab 6 largest CMAs
regions <- list_census_regions(dataset) %>% 
  filter(level=='CMA') %>% 
  top_n(6, pop) %>%
  as_census_region_list
```

### Dynamically Select Vectors

We query the vectors for the *structural type of dwelling* variables and grab all non-summary variables.
```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Search for for the Census vector pertaining to "structural type of dwelling",
# grab all child nodes (leaves only)
# and extract the vectors
vectors <- search_census_vectors("structural type of dwelling",dataset) %>%
  child_census_vectors(TRUE) %>% 
  pull(vector)
```


### Get the census data

Next we grab the census data using the vectors we've selected. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# get data for these CMAs, census variables only (no geographical data)
census_data <- get_census(dataset=dataset, regions=regions, vectors=vectors, level='Regions',geo_format=NA,labels='short')

# get labels using the built in function census_vectors()
categories <- census_vectors(census_data)

# shorten label names where appropriate
categories$Detail <- gsub(" in a building that has ",", ",categories$Detail)

# Rename census variables with the more detailed names above
names(census_data)[grep("v_", names(census_data))] <- categories$Detail
```


### Graph

All that's left to do is graph the data. In this instance, we're putting together a stacked bar chart that shows us a quick comparison of dwelling type across selected Census Metropolitan Areas. To make things look nicer, we first order the dwelling types by their overall prevalence in the sample and the areas by the prevalence of the largest dwelling type.

```{r}
# order dwelling types by prevalence
l <- setNames(lapply(categories$Detail,function(x){sum(census_data[[x]]/census_data$Dwellings)}),categories$Detail)
cats <- names(rev(l[order(unlist(l))]))

# Building types are only for occupied units, add in unoccupied to base out of all dwellings
census_data$`Unoccupied (by usual residents)` <- census_data$Dwellings-census_data$Households
cat_list <- c(cats,"Unoccupied (by usual residents)")

# sort regions by most prevalent dwelling type
region_list <- census_data %>% arrange(UQ(as.name(cat_list[1]))/Dwellings) %>% pull("Region Name")
```


```{r, fig.width=7}

# We want to use ggplot to plot this, and ggplot requires data in a long format. Reshape and gather data using tidyr::gather
plot_data <- census_data %>% tidyr::gather(key = `Dwelling Type`, value = Count, cat_list)

# set correct order on factors 
plot_data$`Dwelling Type` <- factor(plot_data$`Dwelling Type`,levels=rev(cat_list), ordered = TRUE) 
plot_data$`Region Name` <- factor(plot_data$`Region Name`,levels=rev(region_list), ordered=TRUE)


library(ggplot2)

ggplot(plot_data %>% arrange(`Region Name`), aes(x = `Region Name`, y = Count, group = `Dwelling Type`, fill = `Dwelling Type`)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_y_continuous(labels = percent) + 
  labs(x = "8 largest Municipalities in Toronto CMA", y = "Share of dwelling units (%)", title="Dwellings by Structural Type of Dwelling",caption="Canada Census 2016 via cancensus & censusmapper.ca") +
  scale_fill_brewer(palette = "Paired") +
  theme_bw()
```
