---
title: "Bar Graph Demo"
author: "Jens von Bergmann"
date: '2017-08-17'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

We demonstrate how to import data from the [CensusMapper](https://CensusMapper.ca) API. The code for this notbook can be found on [GitHub](https://github.com/mountainMath/cancensus).

This example demonstrates how to dynamically discover regions and vectors and produce an overview graph based on the data by using the `list_census_regions`, `list_census_vectors` and `search_census_vectors` methods.

```{r API Key, echo=TRUE, warning=FALSE}
library(cancensus)
# set your API key
# options(cancensus.api_key='your_api_key')

# use 2016 data
dataset <- 'CA16'
```

### Dynamically Select Regions

We select the top 6 CMAs by population.
```{r, echo=TRUE, message=FALSE, warning=FALSE}
# grab 6 largest CMAs
regions <- list_census_regions(dataset) %>% 
  filter(level=='CMA') %>% 
  top_n(6, pop) %>%
  as_census_region_list
```

### Dynamically Select Vectors

We query the vectors for the *structural type of dwelling* variables and grab all non-summary variables.
```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Search for for the Census vector pertaining to "structural type of dwelling",
# grab all child nodes (leaves only)
# and extract the vectors
vectors <- search_census_vectors("structural type of dwelling",dataset) %>%
  child_census_vectors(TRUE) %>% 
  pull(vector)
```


### Get the census data

Next we grab the census data using the vectors we've selected. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# get data for these CMAs, census variables only (no geographical data)
census_data <- get_census(dataset=dataset, regions=regions, vectors=vectors, level='Regions',geo_format=NA,labels='short')

# get labels using the built in function census_vectors()
categories <- census_vectors(census_data)

# shorten label names where appropriate
categories$Detail <- gsub(" in a building that has ",", ",categories$Detail)

# Rename census variables with the more detailed names above
names(census_data)[grep("v_", names(census_data))] <- categories$Detail

```


### Graph

All that's left to do is graph the data. In this instance, we're putting together a stacked bar chart that shows us a quick comparison of dwelling type across selected Census Metropolitan Areas. 

```{r, fig.width=10}

# We want to use ggplot to plot this, and ggplot requires data in a long format. Reshape and gather data using tidyr::gather
plot_data <- census_data %>% tidyr::gather(key = `Dwelling Type`, value = Count, categories$Detail)

library(ggplot2)
library(scales)

ggplot(plot_data, aes(x = `Region Name`, y = Count, group = `Dwelling Type`, fill = `Dwelling Type`)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_y_continuous(labels = percent) + 
  labs(x = "6 largest CMAs", y = "Share of occupied dwelling units (%)", title="Occupied Private Dwellings by Structural Type of Dwelling",caption="Canada Census 2016 via cancensus & censusmapper.ca") +
  scale_fill_brewer(palette = "Paired") +
  theme_bw()
```
