---
title: "Bar Graph Demo"
author: "Jens von Bergmann"
date: '2017-08-17'
output:
  html_document: default
  html_notebook: default
vignette: |
  %\VignetteIndexEntry{Vignette Title} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

We demonstrate how to import data from the [CensusMapper](https://CensusMapper.ca) API. The code for this notbook can be found on [GitHub](https://github.com/mountainMath/cancensus).

This example demonstrates how to dynamically discover regions and vectors and produce an overview graph based on the data by using the `cancensus.list_regions`, `cancensus.list_vectors` and `cancensus.search_vectors` methods.

```{r API Key, echo=TRUE, warning=FALSE}
library(cancensus)
# set your API key
# options(cancensus.api_key='your_api_key')

# use 2016 data
dataset <- 'CA16'
```

### Dynamically Select Regions

We select the top 6 CMAs by population.
```{r, echo=TRUE, message=FALSE, warning=FALSE}
# grab 6 largest CMAs
cma_list <- cancensus.list_regions(dataset) %>% 
  filter(level=='CMA') %>% 
  top_n(6, pop) %>%
  cancensus.as_region_list
```

### Dynamically Select Vectors

We query the vectors for the *structural type of dwelling* variables and grab all non-summary variables.
```{r, echo=TRUE, message=FALSE, warning=FALSE}

# Search for all Census vectors containing the term "dwellings"
dwellings <- cancensus.search_vectors("structural type of dwelling",dataset)

# Select all dependent variables
dwellings <- cancensus.list_vectors(dataset) %>% filter(parent_vector == dwellings$vector)

# Append all leaves of dependent variables
dwellings <- rbind(dwellings, cancensus.list_vectors(dataset) %>% filter(parent_vector %in% dwellings$vector))

# Keep only leaves
dwellings <- dwellings %>% filter(!vector %in% parent_vector)

# Extract string of vectors we want 
vectors <- dwellings %>% pull(vector)

```


### Get the census data

Next we grab the census data using the vectors we've selected. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# get data for these CMAs
census_data <- cancensus.load(dataset='CA16', regions=regions, vectors=vectors, level='Regions',geo_format=NA,labels='short')

# get labels using the built in function cancensus.labels()

categories <- cancensus.labels(census_data)

# shorten label names where appropriate
categories$Detail <- gsub(" in a building that has ",", ",categories$Detail)

# Rename census variables with the more detailed names above
names(census_data)[grep("v_", names(census_data))] <- categories$Detail

```


### Graph

All that's left to do is graph the data. In this instance, we're putting together a stacked bar chart that shows us a quick comparison of dwelling type across selected Census Metropolitan Areas. 

```{r, fig.width=10}

# We want to use ggplot to plot this, and ggplot requires data in a long format. Reshape and gather data using tidyr::gather
plot_data <- census_data %>% tidyr::gather(key = `Dwelling Type`, value = Count, categories$Detail)

library(ggplot2)
library(scales)

ggplot(plot_data, aes(x = `Region Name`, y = Count, group = `Dwelling Type`, fill = `Dwelling Type`)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_y_continuous(labels = percent) + 
  labs(x = "6 largest CMAs", y = "Share of occupied dwelling units (%)", title="Occupied Private Dwellings by Structural Type of Dwelling",caption="Canada Census 2016 via cancensus & censusmapper.ca") +
  scale_fill_brewer(palette = "Paired") +
  theme_bw()
```
