---
title: "cancensus"
author: "" 
date: ""
output: rmarkdown::html_vignette
mainfont: Roboto

vignette: >
  %\VignetteIndexEntry{cancensus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = nzchar(Sys.getenv("COMPILE_VIG"))
)
library(cancensus)
library(dplyr)
# set_api_key(<your_api_key>, install = TRUE)
```
# Cancensus and CensusMapper

The **cancensus** package was developed to provide users with a way to access Canadian Census in a programmatic way following good [tidy data](http://vita.had.co.nz/papers/tidy-data.pdf) practices. While the structure and data in **cancensus** is unique to Canadian Census data, this package is inspired in part by [tidycensus](https://github.com/walkerke/tidycensus), a package to interface with the US Census Bureau data APIs. 

As Statistics Canada does not provide direct API access to Census data, **cancensus** retrieves Census data indirectly through the [CensusMapper](https://censusmapper.ca) API. CensusMapper is a project by [Jens von Bergmann](https://github.com/mountainMath), one of the authors of **cancensus**, to provide interactive geographic visualizations of Canadian Census data. CensusMapper databases store all publicly available data from Statistics Canada for the 2006, 2011, and 2016 Censuses. Censusmapper data can be accessed via an API and **cancensus** is built to interface directly with it. 

## API Key

**cancensus** requires a valid CensusMapper API key to use. You can obtain a free API key by [signing up](https://censusmapper.ca/users/sign_up) for a CensusMapper account. CensusMapper API keys are free and public API quotas are generous; however, due to incremental costs of serving large quantities of data, there limits to API usage in place. For most use cases, these API limits should not be an issue. Production uses with large extracts of fine grained geographies may run into API quota limits. For larger quotas, please get in touch with Jens [directly](mailto:jens@censusmapper.ca).  

To check your API key, just go to "Edit Profile" (in the top-right of the CensusMapper menu bar). Once you have your key, you can store it in your system environment so it is automatically used in API calls. To do so just enter `set_api_key(<your_api_key>, install = TRUE)`

## Installing cancensus

The stable version of **cancensus** can be easily installed from CRAN.
```{r load_package_cran, echo=TRUE, message=FALSE, warning=FALSE, eval = FALSE}
install.packages("cancensus")

library(cancensus)

options(cancensus.api_key = "your_api_key")
options(cancensus.cache_path = "custom cache path")
```

Alternatively, the latest development version can be installed from Github using `devtools`. 
```{r load_package_git, echo=TRUE, message=FALSE, warning=FALSE, eval = FALSE}
# install.packages("devtools")
devtools::install_github("mountainmath/cancensus")

library(cancensus)

options(cancensus.api_key = "your_api_key")
options(cancensus.cache_path = "custom cache path")
```

For performance reasons, and to avoid unnecessarily drawing down API quotas, **cancensus** caches data queries under the hood. By default, **cancensus** caches in R's temporary directory, but this cache is not persistent across sessions. In order to speed up performance, reduce quota usage, and reduce the need for unnecessary network calls, we recommend assigning a persistent local cache using `set_cache_path(<local cache path>, install = TRUE)`, this enables more efficient loading and reuse of downloaded data.. Users will be prompted with a suggestion to change their default cache location when making API calls if one has not been set yet. 

# Accessing Census Data

**cancensus** provides three different functions for retrieving Census data:
* `get_census` to retrieve Census data and geography as a spatial dataset
* `get_census_data` to retrieve Census data only as a flat data frame
* `get_census_geometry` to retrieve Census geography only as a collection of spatial polygons.

`get_census` takes as inputs a dataset parameter, a list of specified regions, a vector of Census variables, and a Census geography level. You can specify one of three options for spatial formats: `NA` to return data only, `sf` to return an sf-class data frame, or `sp` to return a SpatialPolygonsDataFrame object. 

```{r get_census example, echo=TRUE, warning=FALSE, message=FALSE}
# Returns a data frame with data only
census_data <- get_census(dataset='CA21', regions=list(CMA="59933"),
                          vectors=c("v_CA21_434","v_CA21_435","v_CA21_440"),
                          level='CSD', use_cache = FALSE, geo_format = NA, quiet = TRUE)

# Returns data and geography as an sf-class data frame
census_data <- get_census(dataset='CA21', regions=list(CMA="59933"),
                          vectors=c("v_CA21_434","v_CA21_435","v_CA21_440"),
                          level='CSD', use_cache = FALSE, geo_format = 'sf', quiet = TRUE)

# Returns a SpatialPolygonsDataFrame object with data and geography
census_data <- get_census(dataset='CA21', regions=list(CMA="59933"),
                          vectors=c("v_CA21_434","v_CA21_435","v_CA21_440"),
                          level='CSD', use_cache = FALSE, geo_format = 'sp', quiet = TRUE)
```

**cancensus** utilizes caching to increase speed, minimize API token usage, and to make data available offline. Downloaded data is hashed and stored locally so if a call is made to access the same data, **cancensus** will read the local version instead. To force **cancensus** to refresh the data, specify `use_cache = FALSE` as a parameter for `get_census`. 

Additional parameters for advanced options can be viewed by running `?get_census`.

## Census Datasets

**cancensus** can access Statistics Canada Census data for Census years 1996, 2001, 2006, 2011, and 2016 . You can run `list_census_datasets` to check what datasets are currently available for access through the CensusMapper API. Additional data for the 2016 Census will be included in Censusmapper within a day or two after public release by Statistics Canada. Statistics Canada maintains a release schedule for the Census 2016 Program which can be viewed on their [website](http://www12.statcan.gc.ca/census-recensement/2016/ref/release-dates-diffusion-eng.cfm).

Thanks to contributions by the Canada Mortgage and Housing Corporation (CMHC), **cancensus** now includes additional Census-linked datasets as open-data releases. These include annual taxfiler data at the census tract level for tax years 2000 through 2017, which includes data on incomes and demographics, as well as specialized crosstabs for Structural type of dwelling by Document type, which details occupancy status for residences. These crosstabs are available for the 2001, 2006, 2011, and 2016 Census years at all levels starting with census tract.

The function `list_census_datasets()` will show all available datasets alongside their metadata.

```{r list datasets, message=FALSE, warning=FALSE}
list_census_datasets()
```

As other Census datasets become available via the CensusMapper API, they will be listed as output when calling `list_census_datasets()`. 

## Census Regions

Census data is aggregated at multiple [geographic levels](### Census Geographic Levels). Census geographies at the national (C), provincial (PR), census metropolitan area (CMA), census agglomeration (CA), census division (CD), and census subdivision (CSD) are defined as named census regions. 

Canadian Census geography can change in between Census periods. **cancensus** provides a function, `list_census_regions(dataset)`, to display all named census regions and their corresponding id for a given census dataset. 

```{r list regions, message=FALSE, warning=FALSE}
list_census_regions("CA21")
```

The `regions` parameter in `get_census` requires as input a list of region id strings that correspond to that regions geoid. You can combine different regions together into region lists.

```{r, message=FALSE, warning=FALSE}
# Retrieves Vancouver and Toronto
list_census_regions('CA21') %>% 
  filter(level == "CMA", name %in% c("Vancouver","Toronto"))

census_data <- get_census(dataset='CA21', regions=list(CMA=c("59933","35535")),
                          vectors=c("v_CA21_434","v_CA21_435","v_CA21_440"),
                          level='CSD', use_cache = FALSE, quiet = TRUE)
```

## Census Geographic Levels

Census data accessible through **cancensus** comes is available in a number of different aggregation levels including: 

| Code    | Description                    | Count in Census 2016 |
| --------|--------------------------------|:--------------------:|
| C       | Canada (total)                 | 1                    |
| PR      | Provinces/Territories          | 13                   |
| CMA     | Census Metropolitan Area       | 35                   |
| CA      | Census Agglomeration           | 14                   |
| CD      | Census Division                | 287                  |
| CSD     | Census Subdivision             | 713                  |
| CT      | Census Tracts                  | 5621                 |
| DA      | Dissemination Area             | 56589                |
| EA      | Enumeration Area (1996 only)   | -                    |
| DB      | Dissemination Block (2001-2016)| 489676               |
| Regions | Named Census Region            |                      |

Selecting `regions = "59933"` and `level = "CT"` will return data for all 478 census tracts in the Vancouver Census Metropolitan Area. Selecting `level = "DA"` will return data for all 3450 dissemination areas and selecting `level = "DB"` will retrieve data for 15,197 dissemination block. Working with CT, DA, EA, and especially DB level data significantly increases the size of data downloads and API usage. To help minimize additional overhead, **cancensus** supports local data caching to reduce usage and load times. 

Setting `level = "Regions"` will produce data strictly for the selected region without any tiling of data at lower census aggregation levels levels. 

# Working with Census Variables

Census data contains thousands of different geographic regions as well as thousands of unique variables. In addition to enabling programmatic and reproducible access to Census data, **cancensus** has a number of tools to help users find the data they are looking for. 

## Displaying available Census variables

Run `list_census_vectors(dataset)` to view all available Census variables for a given dataset. 

```{r list_vectors, message=FALSE, warning=FALSE}
list_census_vectors("CA21")
```

## Variable characteristics

For each variable (vector) in that Census dataset, this shows:

* Vector: short variable code 
* Type: variables are provided as aggregates of female responses, male responses, or total (male+female) responses
* Label: detailed variable name
* Units: provides information about whether the variable represents a count integer, a ratio, a percentage, or a currency figure
* Parent_vector: shows the immediate hierarchical parent category for that variable, where appropriate
* Aggregation: indicates how the variable should be aggregated with others, whether it is additive or if it is an average of another variable
* Description: a rough description of a variable based on its hierarchical structure. This is constructed by **cancensus** by recursively traversing the labels for every variable's hierarchy, and facilitates searching for specific variables using key terms. 

## Variable search

Each Census dataset features numerous variables making it a bit of a challenge to find the exact variable you are looking for. There is a function, `find_census_vectors()`, for searching through Census variable metadata in a few different ways. There are three types of searches possible using this function: exact search, which simply looks for exact string matches for a given query against the vector dataset; keyword search, which breaks vector metadata into unigram tokens and then tries to find the vectors with the greatest number of unique matches; and, semantic search which works better with search phrases and has tolerance for inexact searches. Switching between search modes is done using the `query_type` argument when calling `find_census_variables()` function. 

```{r search_vectors1, message=FALSE, warning=FALSE}
# Find the variable indicating the number of people of Austrian ethnic origin
find_census_vectors("Australia", dataset = "CA16", type = "total", query_type = "exact")

find_census_vectors("Australia origin", dataset = "CA16", type = "total", query_type = "semantic")

find_census_vectors("Australian ethnic", dataset = "CA16", type = "total", query_type = "keyword", interactive = FALSE)
```

## Managing variable hierarchy 

Census variables are frequently hierarchical. As an example, consider the variable for the number of people of Austrian ethnic background. We can select that vector and quickly look up its entire hierarchy using `parent_census_vectors` on a vector list.  

```{r parent_vectors, message=FALSE, warning=FALSE}
list_census_vectors("CA16") %>% 
  filter(vector == "v_CA16_4092") %>% 
  parent_census_vectors()
```

Sometimes we want to traverse the hierarchy in the opposite direction. This is frequently required when looking to compare different variable stems that share the same aggregate variable. As an example, if we want to look the total count of Northern European ethnic origin respondents disaggregated by individual countries, it is pretty easy to do so. 

```{r child_vectors1, message=FALSE, warning=FALSE}
# Find the variable indicating the Northern European aggregate
find_census_vectors("Northern European", dataset = "CA16", type = "Total")
```

The search result shows that the vector **v_CA16_4092** represents the aggregate for all Northern European origins. The `child_census_vectors` function can return a list of its constituent underlying variables.

```{r child_vectors2, message=FALSE, warning=FALSE}
# Show all child variable leaves
list_census_vectors("CA16") %>% 
  filter(vector == "v_CA16_4122") %>% child_census_vectors(leaves = TRUE)
```

The `leaves = TRUE` parameter specifies whether intermediate aggregates are included or not. If `TRUE` then only the lowest level variables are returns - the "leaves" of the hierarchical tree. 
