---
title: "Dot-Density Example"
author: "Jens von Bergmann"
date: "2017-08-04"
output: html_notebook
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

First we install cancensus, grab the newest version just in case. Don't forget to set your API key, refer to the [README file](https://github.com/mountainMath/cancensus) for instructions how to get your own.
```{r, echo=TRUE, message=FALSE, warning=FALSE}
#library(devtools)
#devtools::install_github('mountainmath/cancensus')
library(cancensus)
# options(cancensus.api_key)='<your API key>'
```

Load data we want to use for dot density map, use the [CensusMapper API tool](https://censusmapper.ca/api) to select variabes and regions, and we add the `format='sp'` option to read the data in `sp` format.
```{r}
sp <- get_census(geo_format='sp',dataset='CA16', regions=list(CMA="59933"), vectors=c("v_CA16_2066","v_CA16_2060","v_CA16_1916","v_CA16_1727","v_CA16_1937","v_CA16_1958","v_CA16_1973"), level='DA')
```

For plotting, we also want to plot the administrative boundaries as a base layer for better orientation.
```{r}
# load data for background polygons (geograhic data only)
base_geom <- get_census(geo_format ='sp', dataset='CA16', regions=list(CMA="59933"),  level='CSD')
```

Next we neet to prepare the data for mapping, we rename the variables and set the colours
```{r}

#categories to map and their colors
categories=c("Cantonese","Mandarin","Punjabi","Farsi","Tagalog","Korean","Spanish")
colors <- c("#ffff66", "#ff3300","#66ff33","#00ffff","#0066ff","#ff66ff","#6600cc")

# rename columns for better readability
for (i in 1:length(categories)) {x=categories[i]; colnames(sp@data)[grep(x,colnames(sp@data))]<-x}
```


To make the map we have defined a convenience function to handle the dot-density logic. Sparing the details, we need to supply, the data, categories (a vector with the fields we want to map), the colours for each field, a scale (how many people per dot on the map), the base geometry for reference, a map title and map attribution.
```{r dot-density function, include=FALSE}
# a couple of libraries we will need for the dot-density map
library(maptools)
library(rgeos)
library(dplyr)
library(ggplot2)
library(rgdal)
library(ggthemes)

#` Dot density plot function
#`
#` inspired by https://www.blog.cultureofinsight.com/2017/06/building-dot-density-maps-with-uk-census-data-in-r/
#`
#` example
#` dot_density(sp,categories,colors,scale,base_geom,'Chinese spoken at Home')
dot_density <- function(geo_data,categories,colors,scale,base_geo_data,title,attribution){

  # non-random rounding
  num.dots <-  select(geo_data@data, categories) / scale  
  num.dots[is.na(num.dots) | num.dots < 0] <- 0 # clean up NAs and rounding issues

  # do random rounding so we don't loose dots that are spread thinly
  # should be optimized
  random_round <- function(x) {
   v=as.integer(num.dots[, x])
   r=num.dots[, x]-v
   test=runif(length(r), 0.0, 1.0)
   add=rep(0,length(r))
   add[r>test] <- 1
   value=v+add
   return(v)
 }

  geo_data.dfs <- lapply(names(num.dots), function(x) {
    dotsInPolys(sp, random_round(x), f="random")
  })
  
  
  dfs <- lapply(geo_data.dfs, function(x) {
    data.frame(coordinates(x)[,1:2])
  })
  for (i in 1:length(categories)) {
    dfs[[i]]$Category <- categories[i]
  }
  
  # all colors in single vector and randomize order
  
  dots.final <- bind_rows(dfs)
  dots.final$Category <- factor(dots.final$Category, levels = categories)
  
  # randomize order so as not to draw one color at a time, with last color on top.
  dots.random <- dots.final %>% sample_n(length(dots.final$Category))
  
  subtitle=paste('1 dot = ',scale,'people',sep=' ')
  complete_title=paste(title,subtitle,sep="\n")
  complete_attribution=paste("Source:",attribution,"via cancensus & CensusMapper.ca")
  
  # and plot!
  ggplot(base_geo_data) +
    geom_polygon(aes(long, lat, group = group), colour = "black") +
    geom_path(aes(long, lat, group = group), colour = "#d3d3d3", size=0.1) +
    geom_point(data = dots.final, aes(x, y, colour = Category),size=0.1,alpha=0.7) +
    scale_colour_manual(complete_title,values = colors) +
    theme_map() +
    theme(legend.title=element_text(size=18),legend.text=element_text(size=18),plot.background = element_rect(fill = "#eeeeee"), legend.position = "top") +
    guides(colour = guide_legend(override.aes = list(size=3))) +
    labs(color = "label",caption=complete_attribution) +
    coord_map()
  
}
```

All that's left is to call the plot function with our data.

```{r, fig.height=7, fig.width=7}
dot_density(geo_data=sp,categories=categories,colors=colors,scale=25,base_geo_data=base_geom,'Most Common Non-Official Languages at Home',attribution="StatCan Census 2016")
```

