---
title: "Additional datasets: Structural type of dwelling by document type"
output: rmarkdown::html_vignette
mainfont: Roboto
vignette: >
  %\VignetteIndexEntry{Dwellings_by_document_type_cross_tabulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = nzchar(Sys.getenv("COMPILE_VIG"))
)
```

## Background

Through collaboration with the [Canadian Mortgage and Housing Corporation](https://www.cmhc-schl.gc.ca/) (CMHC), CensusMapper has added and open-sourced a special cross-tabulation with *Structural Type of Dwelling* by *Document Type* down to the Census Tract level for the census years 2001, 2006, 2011 and 2016. *Structural Type of Dwelling* is a common census variable that describes the type of structure a dwelling unit is in. *Document Type* is a less frequently used variable that classifies whether the census determined the dwelling is either:

* occupied by usual residents (also known as a household);
* occupied by temporarily present persons; or,
* unoccupied.

This cross-tabulation has information on the structural type of the entire building stock, not just the occupied dwelling units. This is useful when trying to understand the built-up fabric of urban environments.

As an example, we look at the structure of the dwelling stock in the City of Toronto in 2016.

## Example usage: buildings unoccupied vs not occupied by usual residents

Dwellings registered as *unoccupied* on Census day capture the imagination of many, although people often mistakenly pull data on dwellings *not occupied by usual residents* as it is easily available in the standard Census profile data. The advantage of this custom cross-tabulation is that it allows researchers to zoom in on dwellings that were classified as *unoccupied* by the enumerator on Census day for additional detail. 
```{r setup, message = FALSE, warning = FALSE}
# Packages used in this example
library(cancensus)
library(dplyr)
library(tidyr)
library(ggplot2)
```
In this example, we want to retrieve the custom structural dwelling cross-tab for the 2016 Census year with the code `CA16xSD` for the Toronto Census subdivision with the standard Statistics Canada region code `3520005`. For more background on searching for Census geographical regions, see `?list_census_regions()` or the [Get started with cancensus vignette](https://mountainmath.github.io/cancensus/articles/cancensus.html).
```{r, message = FALSE, warning = FALSE}
# Select all variables matching this pattern - 
# Jens, can you explain what this regex is doing? 
vars <- list_census_vectors("CA16xSD") %>% 
  filter(grepl("_\\d$",vector))
variables <- setNames(vars$vector,vars$label)

# Simplify the code below. It's too concise for a vignette. 
census_data <- get_census("CA16xSD",regions=list(CSD="3520005"), vectors = variables, quiet = TRUE) %>%
  pivot_longer(setdiff(names(variables),"Total dwellings")) %>%
  mutate(share=value/sum(value))
```

To visualize what this looks like on a bar chart:
```{r, message = FALSE, warning = FALSE}
ggplot(census_data,aes(x=reorder(name,share),y=share)) +
  geom_bar(stat="identity",fill="steelblue") +
  coord_flip() +
  scale_y_continuous(labels=scales::percent) +
  labs(title="City of Toronto dwelling units by structural type",
       x=NULL,y=NULL,caption="StatCan 2016 Census, xtab via CMHC")
```

As with regular Census data, all data can be retrieved as spatial data. 
```{r, message = FALSE, warning = FALSE}
# Select relevant variables
vars <- list_census_vectors("CA16xSD") %>% 
  filter(grepl("_\\d$",vector))
variables <- setNames(vars$vector,vars$label)

# Retrieve data with attached geography
census_data <- get_census("CA16xSD",regions=list(CSD="3520005"), level="CT", quiet = TRUE, geo_format = "sf",
                          vectors = c(Total="v_CA16xSD_1", Unoccupied="v_CA16xSD_28"),use_cache = FALSE) %>%
  mutate(share=Unoccupied/Total)

# Visualize
ggplot(census_data,aes(fill=share)) +
  geom_sf(size=0.1) +
  scale_fill_viridis_c(labels=scales::percent) +
  coord_sf(datum=NA) +
  labs(title="City of Toronto dwellings unoccupied on census day",
       fill="Share",
       x=NULL,y=NULL,caption="StatCan 2016 Census, xtab via CMHC")
```
