---
title: "Finding intersecting geometries from custom data"
subtitle: "Bring custom data and identify corresponding census geographies"
output: rmarkdown::html_vignette
mainfont: Roboto
vignette: >
  %\VignetteIndexEntry{Finding intersecting geometries from custom data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE, 
  warning = FALSE,
  comment = "#>",
  eval = nzchar(Sys.getenv("COMPILE_VIG"))
)
```

A frequent application of census data is to evaluate values for specific areas that do not necessarily correspond to existing boundaries or administrative units. Census users may have their own defined geographies or other geospatial data of interest and want to be able to quickly and easily identify the collection of census features that correspond to that region.

## A simple example

As an example, suppose we are interested in understanding the housing tenure split in census tracts located near Vancouver Skytrain (rapid transit) stations. We can use the `COV_SKYTRAIN_STATIONS` dataset that ships with the package and is derived from the [City of Vancouver Open Data portal](https://opendata.vancouver.ca/explore/dataset/rapid-transit-stations/information/) and contains their locations. For our example we are interested in census tracts within 800m of these stations, which ships with the package.

```{r setup}
library(cancensus)
library(dplyr)
library(ggplot2)
library(sf)
```

```{r}
cov_station_buffers <- COV_SKYTRAIN_STATIONS %>%
  st_set_crs(4326)  # needed for Ubuntu or systems with old GDAL
```

We then use the `get_intersecting_geometries` call to obtain the list of municipalities (CSDs) and census tracts (CTs) that intersect the 800m station buffer objects. 

```{r}
station_city_ids <- get_intersecting_geometries("CA16", level = "CSD", geometry = cov_station_buffers)
station_ct_ids <- get_intersecting_geometries("CA16", level = "CT", geometry = cov_station_buffers)
```

These return a list of census geographic identifiers suitable for use in the 'region' argument in `get_census`. We may be interested in the transit to work mode share in each of these buffers.

```{r}
variables <- c(mode_base="v_CA16_5792",transit="v_CA16_5801",walk="v_CA16_5804")
station_city <-  get_census("CA16", regions = station_city_ids, vectors = variables, geo_format = 'sf') %>% 
  filter(name == "Vancouver (CY)")
station_cts <-   get_census("CA16", regions = station_ct_ids, vectors = variables, geo_format = 'sf')
```

To understand how these relate we plot the data.

```{r}
ggplot(station_city) +
  geom_sf(fill=NA) +
  geom_sf(data=station_cts,aes(fill=((walk+transit)/mode_base))) +
  geom_sf(data=cov_station_buffers,fill=NA,alpha=0.5,color="steelblue") +
  scale_fill_viridis_c(option="magma",labels=scales::percent) +
  coord_sf(datum=NA) +
  labs(title="CTs within City of Vancouver skytrain station catchments",
       fill="Walk or\ntransit\nto work",
       caption="StatCan Census 2016")
```

To get a closer match we can cut out the dissemination areas intersecting the station catchment areas. 

```{r}
station_das <-   get_intersecting_geometries("CA16", level = "DA", geometry = cov_station_buffers) %>% 
  get_census("CA16", regions = ., vectors=variables, geo_format = 'sf')

ggplot(station_city) +
  geom_sf(fill=NA) +
  geom_sf(data=station_das,aes(fill=((walk+transit)/mode_base))) +
  geom_sf(data=cov_station_buffers,fill=NA,alpha=0.5,color="steelblue") +
  scale_fill_viridis_c(option="magma",labels=scales::percent) +
  coord_sf(datum=NA) +
  labs(title="DAs within City of Vancouver skytrain station catchments",
       fill="Walk or\ntransit\nto work",
       caption="StatCan Census 2016")

```

## Addendum

We may take this further by estimate values of census variables strictly within catchments areas. Rather than intersecting, some adjustments for spatial disaggregation and interpolation are needed. The `tongfen_estimate` method from the [`tongfen` package](https://mountainmath.github.io/tongfen/index.html) is useful in this case. This is a related package that is designed to work in tandem with `cancensus` in order to facilitate census geography aggregation and is designed to make census data comparable across several censuses.

