---
title: "Working with cancensus"
author: "Dmitry Shkolnik"
date: "8/1/2017"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setting up your cancensus environment

```{r, message=FALSE}
# install_github("mountainmath/cancensus")
library(cancensus)
# options(cancensus.api_key)='your_key'
```
## Downloading Census Data

The `get_census` function takes JSON parameters for selecting appropriate census year, the region grouping, the region ids, and a vector of Census variables. 

In order to make this easier, Censusmapper has an [API query playground](https://censusmapper.ca/api/) (requires free account) where you can search for and select specific variables for any given geography within the Canadian census environment. The API playground will generate prepopulated R code for downloading your data, as in the example below: 

```{r, message=FALSE}
census_data <- get_census(dataset='CA11', regions=list(CSD=c("5915022","5915025","5915015","5915011","5915004","5915029","5915046","5915055","5915034")), vectors=c("v_CA11N_2221","v_CA11N_2224","v_CA11N_2227","v_CA11N_2218"), level='CSD', geo_format = 'sf')
```
The data loading functions in **cancensus** are designed to return tidy data frames with matching spatial data. By default, get_census returns geographic data using the sf class, but users can request sp class geographic data by adding the parameter `format="sp"` to `get_census`. For more detail about parameters and options, see `?get_census`. 

## Working with Census data

Because **cancensus** returns tidy data, working with the resulting Census data is straightforward. The following examples rely on `dplyr`-based approaches for manipulating data, but the same tasks can be accomplished in base R and many other approaches. 

```{r, message=FALSE}
library(dplyr)
library(tidyr)

commutes <- census_data %>% 
  group_by(name) %>%
  mutate(share57 = `v_CA11N_2221: Between 5 and 6:59 a.m.` / `v_CA11N_2218: Total employed population aged 15 years and over by time leaving for work`,
         share79 = `v_CA11N_2224: Between 7 and 9:00 a.m.` / `v_CA11N_2218: Total employed population aged 15 years and over by time leaving for work`,
         share9p = `v_CA11N_2227: Anytime after 9:00 a.m.` / `v_CA11N_2218: Total employed population aged 15 years and over by time leaving for work`) %>%
  gather(key = start, value = share, share57:share9p)
```

## Visualizing Census data (except maps)

```{r, message=FALSE}
library(ggplot2)

ggplot(commutes, aes(x = start, y = share)) + geom_bar(stat = "identity") + facet_wrap(~name) + theme_minimal() + scale_x_discrete("",labels = c("share57" = "5-7am", "share79" = "7-9am","share9p" = "After 9am")) + scale_y_continuous("Share of all commuters")+ ggtitle("Share of Commutes by Start Time", subtitle = "2011 NHS, Selected Census Subdivisions")
```

## Mapping Census Data

```{r, message=FALSE}
# Use the latest version of ggplot2 to effortlessly plot sf
# objects and for the viridis colour scales:
# devtools::install_github("tidyverse/ggplot2")
library(ggplot2)

ggplot(commutes) +
  geom_sf(aes(fill = share)) +
  scale_fill_viridis_c("Share of commuters") +
  ggtitle("Commutes by start-time") +
  theme_minimal() + facet_grid(~start) +
  theme(axis.text = element_blank(), 
axis.title = element_blank(), axis.ticks = element_blank()) 
```
